Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019

### 2020-06-22 manuscript/EP1162_manuscript_draft.Rmd
c

## Growth responses to partial cutting
Partial cutting in mixed-species uneven-aged stands requires an understanding of species- and size-specific growth responses to different levels of stand manipulation [@Deal2002].  Our results suggest that a higher level of basal area removal (i.e., a residual basal area (RBA) of 10m^2^ in 1992) stimulated a stronger and more sustained tree-level growth (both volume and basal area increment) response than the high RBA treatment unit and the control unit.  Over each period, basal area increment was higher in the low RBA treatment units compared to other two, and peaked ca. 20 years after treatment, compared to an increment peak of ca. 5 years in high RBA and control units.  At the tree-level, spruce trees grew faster than fir, and in both species, larger trees responded with more growth increment than smaller trees (although this was reversed when increment was expressed as a percentage of original tree size).

@@

@@

Our results show that both species had initial delayed growth response after treatment.  Trees acclimated to shady conditions may suffer a shock after release due to relatively rapid changes in light, temperatures and humidity which can increase evapotranspiration and respiration stress [get sources from @Wang2011].  A post-release lag in height growth response may also reflect the tree's reallocation of of carbon to below-ground structures at the expense of aboveground, as per the functional balance theory [@Kneeshaw2002; @Brouwer1962]; a relatively immediate response in basal area increment response after treatment may represent a species strategy to thicken stem afer release in response to increased wind stress [@Wang2011].

@@

All treatment units maintained a stand structure analagous to an uneven-aged stand (ie reverse-J diameter distribution).  In the low RBA unit, most recruitment was into the smallest size classes, whereas recruitment was into larger size classes in the high RBA and control units.  This suggests that understory response may increase with level of change in the canopy, and that  more substantial changes to the stand conditions can create more favourable conditions for understory suppressed and intermediate trees to release and increase size.  We did not account for trees smaller than 7.5cm DBH in the study, therefore, we do not have data on seedling germination and establishment after treatment.  THe increase of tree density in the lower size classes in the low RBA may have been a result of increased seedling establishment after harvest.  However, our data show that tree density increased in the smaller size classes in the low RBA unit.  Natural regeneration is complex and in some cases can be largely stochastic, which can make its use difficult to achieve management objectives [@Bataineh2013].

@@

@@

>>>>

## Growth responses to partial cutting
Partial cutting in mixed-species uneven-aged stands requires an understanding of species- and size-specific growth responses to different levels of stand manipulation [@Deal2002].  Our results suggest that a higher level of basal area removal (i.e., a residual basal area (RBA) of 10m^2^ in 1992) stimulated a stronger and more sustained tree-level growth (both volume and basal area increment) response than the high RBA treatment unit and the control unit.

Over each period, basal area increment was higher in the low RBA treatment units compared to other two, and peaked ca. 20 years after treatment, compared to an increment peak of ca. 5 years in high RBA and control units.  At the tree-level, spruce trees grew faster than fir, and in both species, larger trees responded with more growth increment than smaller trees (although this was reversed when increment was expressed as a percentage of original tree size).  Longer periods (i.e., over 40 years) of increased growth rates have been documented after understorey release by insect attack in subalpine spruce-fir forests in Colorado, US [@Veblen1991].

@@

Spruce trees in the understorey can have a clumpier distribution than subalpine fir, which may indicate that spruce has a narrower range of suitable microsites for regeneration [@Shea1985; @Klinka2000].  On the other hand, subalpine fir can regenerate across a wider range of substrates, and therefore may be expected to proliferate more uniformly in the understorey [@Klinka2000].   Thus, partial harvesting systems that aim to maintain spruce composition in the stand should aim to protect spruce understorey as much as possible, as well as create suitable microsites for seed germination such as exposed mineral soil [@Coates1994].  Once established, the performance of planted spruce and subalpine may be more or less equal [@references].  Understorey Engelmann spruce and subalpine fir displayed similar growth responses to release after bark beetle attack in a subalpine forests in Colorado [@Veblen1991].

@@

Our results show that both species had initial delayed growth response after treatment.  Trees acclimated to shady conditions may suffer a shock after release due to relatively rapid changes in light, temperatures and humidity which can increase evapotranspiration and respiration stress [get sources from @Wang2011].  A post-release lag in height growth response may also reflect the tree's reallocation of of carbon to below-ground structures at the expense of aboveground, as per the functional balance theory [@Kneeshaw2002; @Brouwer1962], whereas a relatively immediate response in basal area increment response after treatment may represent a species strategy to thicken stem afer release in response to increased wind stress [@Wang2011].

Hybrid spruce and subalpine fir share similar silvics and life histories [@Klinka2000], and have demonstrated similar growth and survival in managed forests in British Columbia [@Lajzerowicz2006] and elsewhere [@Veblen1991].  Thus, we may have anticipated similar growth responses between these species.

@@

All treatment units maintained a stand structure analagous to an uneven-aged stand (ie reverse-J diameter distribution).  In the low RBA unit, most recruitment was into the smallest size classes, whereas recruitment was into larger size classes in the high RBA and control units.  This suggests that understory response may increase with level of change in the canopy, and that  more substantial changes to the stand conditions can create more favourable conditions for understory suppressed and intermediate trees to release and increase size.  We did not account for trees smaller than 7.5cm DBH in the study, therefore, we do not have data on seedling germination and establishment after treatment.  THe increase of tree density in the lower size classes in the low RBA may have been a result of increased seedling establishment after harvest.  However, our data show that tree density increased in the smaller size classes in the low RBA unit.  Natural regeneration is complex and in some cases can be largely stochastic, which can make its use difficult to achieve management objectives [@Bataineh2013].  We note that in high-elevation spruce-fir stands in British Columbia, understorey tree age and diameter may be uncorrelated [@Puttonen1998].

@@

We note that our data do not provide information on the effects of partial harvest removal on spatial distribution of tree stems, which may be an important consideration when applying silvicultural treatments [@Saunders2008].  Spatial characteristics include spatial distribution of tree stems (e.g., clumped or even distribution), the degree of species intermingling, and size differences among neighbours, all of which can influence resource availablility, regeneration success and competition [cf @Saunders2008].

@@

Both treatments resulted in a stand that would be considered fully stocked immediately after harvest according to the BC free growing guidelines [@reference].
